# GettingData-README
#### _Chris Vennes_
#### _August 22, 2015_


This ReadMe contains on executing and creation of run_analysis.R.  

### Running run_analysis.R:
Run this file in the same directory as the output of the zip file below.  It should unzip to a file called UCI HAR Dataset; the script should be run in the same working directory as that folder, not inside the folder.
  [Dataset](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip )

Please note that the analysis script requires dplyr packages to be installed, if you do not have dplyr then instal it:

```{r eval=FALSE}
  install.packages("dplyr")
```

The output of the script will be a table containing Tidy Data as docuemnted by the [CodeBook](https://github.com/cjvennes/DataScience/blob/master/03%20GettingData/Course%20Project/CodeBook.Rmd) in this repository.

### Creation and Walkthrough of run_analysis.R:

See the full script [here](https://github.com/cjvennes/DataScience/blob/master/03%20GettingData/Course%20Project/run_analysis.R) with limited notes embedded.

The function requires no inputs but does require you to run in the same working folder as the UCI HAR Dataset (see above).

The main objectives our of analysis as specified by the course project are:
    1. Merges the training and the test sets to create one data set.
    2. Extracts only the measurements on the mean and standard deviation for each measurement. 
    3. Uses descriptive activity names to name the activities in the data set
    4. Appropriately labels the data set with descriptive variable names. 
    5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

**_STEP 1 _**

_1. Load the data and merge training and test sets to create one data set._ 
First we need to load all applicable datasets we want to use:

```{r eval=FALSE}
    activityLabels <- read.table("./UCI HAR Dataset/activity_labels.txt")
    features <- read.table("./UCI HAR Dataset/features.txt")
    subjectTest <- read.table("./UCI HAR Dataset/test/subject_test.txt")
    subjectTrain <- read.table("./UCI HAR Dataset/train/subject_train.txt")
    Xtest <- read.table("./UCI HAR Dataset/test/X_test.txt")
    Xtrain <- read.table("./UCI HAR Dataset/train/X_train.txt")
    labelTest <- read.table("./UCI HAR Dataset/test/y_test.txt")
    labelTrain <- read.table("./UCI HAR Dataset/train/y_train.txt")
```

For ease of use and save us a more complicate step later we will add headers to the Test and Training tables from the features.txt data:

```{r eval=FALSE}
  names(Xtest) <- features[,2]
  names(Xtrain) <- features[,2]
```

Now we will combine the two datasets into a single table called Xdata

```{r eval=FALSE}
  Xdata <- rbind(Xtest,Xtrain)
```

**_STEP 2 _**

_2. Extract only the variables with mean or standard deviation._

Use grep with the following patter to search for mean and std:
```{r eval=FALSE}
  pattern <- c("-mean()","-std()")
  data <- Xdata[,grep(paste(pattern,collapse="|"), names(Xdata))]
```

There is another meanFreq column that we do not want, but isn't removed with the above script.  So remove that one too
```{r eval=FALSE}
  Xdata <- Xdata[,grep("meanFreq", names(Xdata), invert=TRUE)]
```
    
**_STEP 3_**

_3. Uses descriptive activity names to name the activities in the data set_

Combine the subject tables that we will use to merge to Xdata above
```{r eval=FALSE}
    subjectTest$Set <- "Test"  #This will be used once we merge the two sets to differeniate
    subjectTrain$Set <- "Training" #This will be used once we merge the two sets to differeniate
    subjectX <- rbind(subjectTest, subjectTrain)
    names(subjectX) <- c("Subject","Set")
```

Combine the activity labels that we will use to merge to Xdata above.
```{r eval=FALSE}
    names(activityLabels) <- c("Activity.Code","Activity.Name")
    labelX <- rbind(labelTest, labelTrain)
    names(labelX) <- "Activity.Code"
    labelX <- merge(labelX, activityLabels, by="Activity.Code")
```
Add the Activity and Subject to the main data table, this will apply descriptive labels to the dataset.  Name the output data.

```{r eval=FALSE}
    data <- cbind(labelX, Xdata)
    data <- cbind(subjectX, data)
```  
With this step done we actually have completed objective 4:

_4. Appropriately labels the data set with descriptive variable names._

**_STEP 4_**

_5. From the data set in step 4, create a second, independent tidy data set with the average of each variable for each activity and each subject._

We are going to use the dplyr library to summarize the data.  Please make sure dplyr is installed, the script will load the package if not done so already.
    
create a group by table using the Subject, and Activity Name.  This will allow dpylr to aggregate based on those two variables.
```{r eval=FALSE}
    Group.data <- group_by(data, Subject, Activity.Name)
```  

Summarize data by Subject and Activity.Name, taking the mean of each variable.  Columns 5 through 70 are all variable columns we want to take the mean of.
```{r eval=FALSE}
    Avg.data <- summarise_each(Group.data, funs(mean),5:70)
```      

Finally return this tidy data set for usage in further analysis.
```{r eval=FALSE}
    return(Avg.data)
```
}